Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (ACC with CQN normlaisation)

### 2018-11-07 code/Rmd/CMC_ACC_RSEM_GRCh38_CQN_ModSel.Rmd
Minor changes

---
title: "Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (ACC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)

@@

@@

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "ensembl.org", dataset = "hsapiens_gene_ensembl")#change to specific ensembl archive....host

@@

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = NA
  x[x > (mn+3*std.dev)] = NA
  return(x)
}) %>% t

# Back calculate counts for further analysis
NEW.COUNTS = PROCESSED_COUNTS$filteredExprMatrix$counts
NEW.COUNTS[is.na(LOG.CPM)] = 0
```

@@

# my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
# 
# p = list()
# p[[1]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "X"), aes(x = `Reported Gender`, y = `CPM(log)`)) + geom_boxplot()
# p[[1]] = p[[1]] + ggtitle('X') + my.theme
# p[[2]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y"), aes(x = `Reported Gender`, y = `CPM(log)`)) + geom_boxplot()
# p[[2]] = p[[2]] + ggtitle('Y') + my.theme
#multiplot(plotlist = p, cols = 2)
```
### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
GENE.LEN = tibble::column_to_rownames(GENE.LEN, var = 'gene_id') %>% as.data.frame()
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (ACC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)

@@

library(future)
library(furrr)

plan(multiprocess)

@@

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "oct2018.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")#change to specific ensembl archive....host

@@

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T) %>%
  rownameToFirstColumn('gene_id') %>%
  tidyr::gather(sample_id, value, -gene_id) %>%
  group_by(gene_id) %>%
  nest() %>%
  tibble::deframe() %>%
  future_map(function(x){
    mn = mean(x$value, na.rm = T)
    std.dev = sd(x$value, na.rm = T)

    x[which(x$value <= mn-3*std.dev), 'value'] = NA 
    x[which(x$value >= mn+3*std.dev), 'value'] = NA

    return(x)
  }, .progress = T) %>%
  bind_rows(.id = 'gene_id') %>%
  tidyr::spread(sample_id, value) %>%
  tibble::column_to_rownames(var = 'gene_id') %>%
  as.data.frame()

# Assign zero values to NAs
LOG.CPM = LOG.CPM[rownames(PROCESSED_COUNTS$filteredExprMatrix$counts), colnames(PROCESSED_COUNTS$filteredExprMatrix$counts)]
NEW.COUNTS = PROCESSED_COUNTS$filteredExprMatrix$counts
NEW.COUNTS[is.na(LOG.CPM)] = 0
```

@@

# my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
# 
# p = list()
# p[[1]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "X"), aes(x = `Reported Gender`, y = `CPM(log)`)) + geom_boxplot()
# p[[1]] = p[[1]] + ggtitle('X') + my.theme
# p[[2]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y"), aes(x = `Reported Gender`, y = `CPM(log)`)) + geom_boxplot()
# p[[2]] = p[[2]] + ggtitle('Y') + my.theme
#multiplot(plotlist = p, cols = 2)
```
### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
GENE.LEN = tibble::column_to_rownames(GENE.LEN, var = 'gene_id') %>% as.data.frame()
CQN.GENE_EXPRESSION = cqn(round(NEW.COUNTS[,c(1:190, 192:476)]), 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = TRUE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
